Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-31 paper.qmd
sorted many notes and moved hurricane comment over
revised Intro

@@

## Introduction
From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. Information visualisations remain one of the most powerful tools for fast and reliable science communication.

There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success. Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised. Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data and **how** it might diverge from what we expect.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-anscombe
#| fig-cap: "The four scatter plots that make up Anscombe's quartet. The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation. The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
library(tidyverse)
tibble(x = c(anscombe$x1, anscombe$x2,
             anscombe$x3, anscombe$x4),
       y = c(anscombe$y1, anscombe$y2,
             anscombe$y3, anscombe$y4),
       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
                rep("Plot 3",11), rep("Plot 4",11))) %>%
  ggplot(aes(x,y)) +
  geom_point(aes(fill=Plot), colour="black", 
             size=3, pch=21, alpha=0.75) +
  facet_wrap(~Plot) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  scale_fill_brewer(type = "qual", palette = 4)
```

Uncertainty visualisation is a somewhat new subfield of visualisation. Early papers that specifically reference "uncertainty visualisation" start popping up in the late 80s [@Ibrekk1987], with papers in the early 90s (specifically in the area of geospatial information visualisation) recgonising it as a necessary subfield of visualisation [@MacEachren1992; Carr1992]. Despite the youth of the specific sub-field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves. In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.

## Similar work

Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information. Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews.

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field. @Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.

There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature. Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016]. Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017]. Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement. In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.

@@

Another method used by authors is to ask a deterministic question about a random event. @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions. In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution. Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage. The flip side of this is asking participants for a deterministic answer to a probabilistic question. @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.

>>>>

<!-- TODO 
- convert to American spelling
- should it be data-driven or two words
- May not need to reference the same paper again and again
-->

@@

<!-- ## Introduction-->
From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. Information visualisations remain one of the most powerful tools for fast and reliable science communication.

There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness. (@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised. <!-- Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features: DON'T NEED MUCH EMPHASIS ON THIS DATA - VERY WELL-KNOWN -->
Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: false
#| label: fig-anscombe
#| fig-cap: "The four scatter plots that make up Anscombe's quartet. The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation. The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
library(tidyverse)
tibble(x = c(anscombe$x1, anscombe$x2,
             anscombe$x3, anscombe$x4),
       y = c(anscombe$y1, anscombe$y2,
             anscombe$y3, anscombe$y4),
       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
                rep("Plot 3",11), rep("Plot 4",11))) %>%
  ggplot(aes(x,y)) +
  geom_point(aes(fill=Plot), colour="black", 
             size=3, pch=21, alpha=0.75) +
  facet_wrap(~Plot) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  scale_fill_brewer(type = "qual", palette = 4)
```

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display (@MacEachren1992; @Carr1992). Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. <!--In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.-->

<!-- ## Similar work-->

<!-- Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information. Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews. -->

However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions. With the abundance of publications it is timely to consider a review of the state-of-the-art. In fact, there have already been several reviews published.

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarise the papers from this field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of <!--each uncertainty visualisation--> a collection of papers.

There are several reasons provided to explain the difficulty in generalising <!--the findings in--> the uncertainty visualisation literature. Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016]. Other papers indirectly hint at this <!--Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that--> by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017]. Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These concerns appear <!--difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking--> across all application areas,  <!--, in geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].--> and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.

<!--This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.-->

The review provided here attempts to address these issues. We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller components. The ideal role of this review is to provide a guide for research in these topics, to help synthesize this noisy field. <!-- XXX Its not clear that these last two sentences are helpful here.--> <!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.-->

@@

Another method used by authors is to ask a deterministic question about a random event. @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions. In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution. Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. The flip side of this is asking participants for a deterministic answer to a probabilistic question. @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
